Determinism, Complexity, and Predictability in Computer Performance

### 2014-02-10 pecomplexity.tex
New Figures Outlines and results

\newtheorem*{mydef}{Definition}
\title{Determinism, Complexity, and Predictability in Computer Performance}

@@

Computers are among the most complex engineered artifacts in current
use.  Modern microprocessor chips contain multiple processing units
and multi-layer memories, for instance, and they use complicated
hardware/software strategies to move data and threads of computation
across those resources.  These features---along with all the others
that go into the design of these chips---make the patterns of their
processor loads and memory accesses highly complex and hard to
predict.  Accurate forecasts of these quantities, if one could
construct them, could be used to improve computer design.  If one
could predict that a particular computational thread would be bogged
down for the next 0.6 seconds waiting for data from main memory, for
instance, one could save power by putting that thread on hold for that
time period (e.g., by migrating it to a processing unit whose clock
speed is scaled back).  Computer performance traces are, however, very
complex.  Even a simple ``microkernel,'' like a three-line loop that
repeatedly initializes a matrix in column-major order, can produce
{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
Figure~\ref{fig:col-ts}, and chaos places fundamental limits on
predictability.
%
% \begin{figure}[htbp]
%    \centering
%    \includegraphics[width=0.5\textwidth]{figs/colCacheShort}
%    % where an .eps filename suffix will be assumed under latex,
%    % and a .pdf suffix will be assumed for pdflatex
%    \caption{A small snippet of the L2 cache miss rate of {\tt
%        col\_major}, a three-line C program that repeatedly initializes
%      a matrix in column-major order, running on an Intel Core
%      Duo\textsuperscript{\textregistered}-based machine.  Even this
%      simple program exhibits chaotic performance dynamics.}
%    \label{fig:cache}
%  \end{figure}

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/colshortts}
    \caption{A short segment of the instructions executed per CPU clock cycle
    (IPC) during the execution of \col. Each point is the IPC during a 100,000
    cycle period. There is a great deal of periodicity in this time series.}
    \label{fig:col-ts}
  \end{subfigure}%
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
    \caption{The full IPC time series during the execution of \gcc. This time
    series has very little structure.}
    \label{fig:gcc-ts}
  \end{subfigure}
  \label{fig:sample-ts}
\end{figure}

The computer systems community has applied a variety of prediction
strategies to traces like this, most of which employ regression.  An
appealing alternative builds on the recently established fact that
computers can be effectively modeled as deterministic nonlinear
dynamical systems \cite{mytkowicz09}.  This result implies the
existence of a deterministic forecast rule for those dynamics.  In
particular, one can use \emph{delay-coordinate embedding} to
reconstruct the underlying dynamics of computer performance, then use
the resulting model to forecast the future values of computer
performance metrics such as memory or processor loads
\cite{josh-ida2011}.  In the case of simple microkernels like the one
that produced the trace in Figure~\ref{fig:cache}, this deterministic
modeling and forecast strategy works very well.  In more-complicated
programs, however, such as speech recognition software or compilers,
this forecast strategy---as well as the traditional methods---break
down quickly.

This paper is a first step in understanding when, why, and how
deterministic forecast strategies fail when they are applied to
deterministic systems.  We focus here on the specific example of
computer performance.  We conjecture that the complexity of traces
from these systems---which results from the inherent dimension,
nonlinearity, and nonstationarity of the dynamics, as well as from
measurement issues like noise, aggregation, and finite data
length---can make those deterministic signals \emph{effectively}
unpredictable.  We argue that \emph{permutation entropy}
\cite{bandt2002per}, a method for measuring the entropy of a
real-valued-finite-length time series through ordinal analysis, is an
effective way to explore that conjecture.  We study four
examples---two simple microkernels and two complex programs from the
SPEC benchmark suite---running on different Intel-based machines.  For
each program, we calculate the permutation entropy of the processor
load (instructions per cycle) and memory-use efficiency (cache-miss
rates), then compare that to the prediction accuracy attainable for
that trace using a simple deterministic model.

@@

@@

This is an extremely powerful result: among other things, it means
that one can build a formal model of the full system dynamics without
measuring (or even knowing) every one of its state variables.  This is
the foundation of the modeling approach that is used in this paper.
The first step in the process is to estimate values for the two free
parameters in the delay-coordinate map: the delay $\tau$ and the
dimension $m$.  We follow standard procedures for this, choosing the
first minimum in the average mutual information as an estimate of
$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
method of \cite{KBA92}, with a threshold of 10\%, to estimate $m$.  A
plot of the data from Figure~\ref{fig:cache}, embedded following this
procedure, is shown in Figure~\ref{fig:embedding}.

[[Add a plot of an embedding...]]

@@

Given a nonlinear model of a deterministic dynamical system in the
form of a delay-coordinate embedding like Figure~\ref{fig:embedding},
one can build deterministic forecast algorithms by capturing and
exploiting the geometry of the embedding.  Many techniques have been
developed by the dynamical systems community for this purpose
(e.g.,~\cite{casdagli-eubank92,weigend-book}).  Perhaps the most straightforward
is the ``Lorenz method of analogues'' (LMA), which is essentially
nearest-neighbor prediction in the embedded state
space~\cite{lorenz-analogues}.  Even this simple algorithm---which
builds predictions by finding the nearest neighbor in the embedded
space of the given point, then taking that neighbor's path as the
forecast---works quite well on the trace in Figure~\ref{fig:cache}, as
shown in Figure~\ref{fig:cachePredTS}.
%
\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/colCachePredTS}
    \caption{A forecast of the last 4,000 points of the signal in
      Figure~\ref{fig:cache} using an LMA-based strategy on the
      embedding in Figure~\ref{fig:embedding}.  Red circles and blue
      $\times$s are the true and predicted values, respectively;
      vertical bars show where these values differ. }
\label{fig:cachePredTS}
\end{figure}
%
On the other hand, if we use the same approach to forecast the
processor load\footnote{Instructions per cycle, or IPC} of the {\tt
  482.sphinx3} program from the SPEC cpu2006 benchmark suite, running
on an Intel i7\textsuperscript{\textregistered}-based machine, the
prediction is far less accurate; see Figure~\ref{fig:predsphinx}.
%
\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/sphinxPredicTS}
     \caption{An LMA-based forecast of the last 4,000 points of a
       processor-load performance trace from the {\tt 482.sphinx3}
       benchmark.  Red circles and blue $\times$s are the true and
       predicted values, respectively; vertical bars show where these
       values differ.}
\label{fig:predsphinx}
\end{figure}

@@

The weighted permutation entropy for the {\tt SVD} program is given in
Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
the time series. Within each of those windows, the statistics over words of
length 4 are computed and the WPE is calculated. The gray bands denote regions
where the 5,000 value window overlapped visually-distinct regimes. It can be
seen that the behaviors of the weighted permutation entropy vary between
regimes. [[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series differnet levels of complexity and predictability exist. Maybe point at some of the predicting predictability papers.]]

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
  \caption{[[Joshua: I think adding the colored SVD trace to this would be good or putting it above this figure]]The weighted permutation entropy of one run of SVD. The gray bands
    are regions where the window overlaps regimes. The window size used is
    $5,000 \times 100,000$ instructions and the word length is $4$.}
  \label{fig:wwpe}
\end{figure}

@@

%\item Introduce MASE and explain why random walk isn't *that* naive

%We calculated a figure of merit for each prediction method as follows.

@@

In order to analyze correctness of each prediction we split each time series
into two pieces: the first 90\% refereed to as the ``learning/training" signal,
$\{X_{i,obs}\}_{i=1}^{i=n}$ and the last 10\% known as the ``test/correct"
signal $\{c_j\}_{j=n+1}^{k+n+1}$. The learning signal is used to build the
initial models (e.g., LMA, ARIMA, na\"ive) as described in Section
\ref{sec:compModel}. The test signal is used both to test the forecasts and to
refit the models after forecasting. In particular, using the models we perform
$k$ 1-step predictions $\{p_j\}_{j=n+1}^{k+n+1}$, after each prediction we
append the training signal with the the next point in the correct signal. We
would like to note that this rebuilding occurs due to a problem with ARIMA
models converging to a mean prediction if too long of a prediction horizon is
used, this is not a handicap of either LMA or na\"ive.

In order to compare the resulting forecasts we calculate the Mean Absolute
Squared Error (MASE)\cite{MASE} between the true and predicted signals:

\begin{align*}
  \textrm{MASE} = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}
\end{align*}

The scaling term here:

$$s=\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i,obs}|$$ is the average in-sample
forecast error for a random walk prediction $(p_j=c_{j-1})$. This error method
was introduced in \cite{MASE} as a ``generally applicable measurement of
forecast accuracy without the problems seen in the other measurements." The
major advantage of MASE is that it allows fair comparison across methods,
prediction horizons and varying signal scales. A $MASE<1$ means that the
prediction method gives, on average, smaller errors than the 1-step errors from
the random walk forecast strategy. Analogously, a $MASE>1$ means that the
prediction method did worse, on average than the 1-step errors from the random
walk forecast strategy. In Table \ref{tab:error} we provide the distribution of
MASEs for 15 runs of each of the 8 signals across the 3 prediction forecast
methods. For comparison Table \ref{tab:error} also has weighted permutation

%``The errors have been scaled by the one-step in- sample forecast errors from the na\"ive method, and then averaged across all series. So a value of 2 indicates that the out-of-sample forecast errors are, on average, about twice as large as the in-sample one-step forecast errors from the na\"ive method. Because the scaling is based on one-step forecasts, the scaled errors for multi-step forecasts are typically larger than one. "

\begin{table}[htdp]
\caption{Average MASE for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average WPE at word length 5 and 6 for each signal. }
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve & $l=5$  & $l=6$ \\
\hline
\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $1.7970\pm0.0095$& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\

@@

%This is then used as a scaling term in the following way:

%$$qt = \frac{e_t}{\frac{1}{n-1}\}sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$

%$$MASE = \frac{\frac{\sum_{j=n+1}^{k+n+1} |c_j-p_j| }{k}}{\sum^n_{i=2}|X_{i,obs}-X_{i,obs}|}$$

%and is comprised of the first 90\% of the time series, the learning signal is what we build the initial forecast models with. The second piece is called the test  the  back the last 10\% of each signal. Built the initial forecast model for each signal. We then perform an in-sample random walk forecast at every point in the training

%Possible new error measure

%$$MASE = mean(|q_t|)$$

%Helpful to remember that $F_i = Y_{i-1}$ for random walk prediction

%``The errors have been scaled by the one-step in- sample forecast errors from the na\"ive method, and then averaged across all series. So a value of 2 indicates that the out-of-sample forecast errors are, on average, about twice as large as the in-sample one-step forecast errors from the na\"ive method. Because the scaling is based on one-step forecasts, the scaled errors for multi-step forecasts are typically larger than one. "

%\item Paragraph on how to interpret MASE
\begin{enumerate}

\item First paragraph: WPE is a good measure of predictability.  Figure:
best athlete MASE vs. its WPE.

As can be seen in Table \ref{tab:error} the relationship between prediction
accuracy and the weighted permutation entropy (WPE) is much as we conjectured:
performance traces with high WPE are indeed harder to predict using the
forecasting models described in Section \ref{sec:compModel}.
Figure~\ref{fig:pred_vs_wpe} demonstrates this primary finding, with the
exception of one outlier whose behavior we can explain. In
Figure~\ref{fig:pred_vs_wpe} we plot the best prediction (i.e., the lowest MASE
over methods on that signal) for each of the 8 signals.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
  \caption{The best MASE among all runs and prediction methods vs weighted
  permutation entropy. For each of these, the word length used is $6$. The
  dashed line is a least-squares linear fit of all the points except for {\tt
  SVD$_1$} which we have excluded for reasons explained in the text.}
  \label{fig:pred_vs_wpe}
\end{figure}
We find that the relationship between the two is roughly linear. The single
outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE:
namely that in the absence of any large features, as all the other regimes have,
the WPE effectively falls back to the standard PE which the noisy behavior
drives toward 1.0.

This primary finding: ``Signals that are more complex are harder to predict." At
first glance this finding may seem incredibly trivial, of course a complex
signal is harder to predict than a trivial one, but this is \emph{not} the core
result here as we will elucidate now with Figure \ref{fig:predictions_vs_wpe}.
\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{figs/LMA_prediction_vs_entropy}
    \caption{The MASE of LMA vs weighted permutation entropy. For each of these,
    the word length used is $6$.}
    \label{fig:lma_pred_vs_wpe}
  \end{subfigure}%
  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{figs/ARIMA_prediction_vs_entropy}
    \caption{The MASE of ARIMA vs weighted permutation entropy. For each of
    these, the word length used is $6$.}
    \label{fig:arima_pred_vs_wpe}
  \end{subfigure}
  % \begin{subfigure}{0.5\textwidth}
  %   \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
  %   \caption{The MASE values for LMA against ARIMA. The dashed line is the
  %   identity, delineating the traces for which either LMA or ARIMA performed
  %   better. All traces except those from {\tt SVD$_1$} lie above the line,
  %   indicating that LMA is better suited prediction method for the computer
  %  traces considered.}\label{fig:lma_vs_arima}
  % \end{subfigure}
  \label{fig:predictions_vs_wpe}
\end{figure}

\item Talk about structure analysis. Just because there is forward information
transfer, does not mean that linear predictors can get at this.
\item For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by
side plot.
\item full results.  Image: MASE vs. WPE for both LMA \& ARIMA.  Points to make:
\begin{enumerate}
\item clusters are distributed differently
\item clusters are shaped differently---tight or not
\item clusters move differently between LMA and ARIMA
\item finally, the diagonal line is important. If you're below it, you could do
better.

\end{enumerate}
\end{enumerate}

In Figures~\ref{fig:lma_pred_vs_wpe} and \ref{fig:arima_pred_vs_wpe} we directly
compare the performance of the LMA and ARIMA prediction methods (respectively)
to the value of the weighted permutation entropy for all runs of each program
under consideration. The LMA MASE values are largely similar to those of the
best predictions, primarily because LMA often performed superior to ARIMA and
the naive method. On the other hand, the ARIMA MASE values are largely
uncorrelated with WPE values. The fact that ARIMA is uncorrelated with WPE is in
large part one of the major findings of this work. More specifically, say we
tried to predict an arbitrary noisy real-valued time series with an ``out-of-
the-box" prediction strategy like ARIMA as proposed in \cite{ARIMA-autofit}, and
say you got inconsistent and bad forecasts, (i.e., perform worse than the
na\"ive random walk strategy (MASE$>1$). How do we determine if the prediction
strategy is not adequate for the prediction task, or if the signal is simply too
complex to predict. If a signal is too complex and little forward information
transfer is present we may simply not be possible to do better than the random
walk, in which case we should not worry ourselves over finding a more
complicated prediction strategy. However if we measure the complexity to be low,
$\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}), we can likely do better
than the random walk and should search for more adequate prediction strategies.

>>>>

\newtheorem*{mydef}{Definition}
\title{Determinism, Complexity, and Predictability in Computer Performance\\
Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
\\On Quantifying Predictability through Structural Complexity
\\On Quantifying Predictability with Structural Complexity
}

@@

Computers are among the most complex engineered artifacts in current
use.  Modern microprocessor chips contain multiple processing units
and multi-layer memories, for instance, and they use complicated
hardware/software strategies to move data and threads of computation
across those resources.  These features---along with all the others
that go into the design of these chips---make the patterns of their
processor loads and memory accesses highly complex and hard to
predict.  Accurate forecasts of these quantities, if one could
construct them, could be used to improve computer design.  If one
could predict that a particular computational thread would be bogged
down for the next 0.6 seconds waiting for data from main memory, for
instance, one could save power by putting that thread on hold for that
time period (e.g., by migrating it to a processing unit whose clock
speed is scaled back).  Computer performance traces are, however, very
complex.  Even a simple ``microkernel,'' like a three-line loop that
repeatedly initializes a matrix in column-major order, can produce
{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
Figure~\ref{fig:ipc}, and chaos places fundamental limits on
predictability.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A small snippet of the instructions per cycle(ipc) of {\tt
        \col}, a three-line C program that repeatedly initializes
      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.  Even this
      simple program exhibits chaotic performance dynamics.}
   \label{fig:ipc}
  \end{figure}

The computer systems community has applied a variety of prediction
strategies to traces like this, most of which employ regression.  An
appealing alternative builds on the recently established fact that
computers can be effectively modeled as deterministic nonlinear
dynamical systems \cite{mytkowicz09}.  This result implies the
existence of a deterministic forecast rule for those dynamics.  In
particular, one can use \emph{delay-coordinate embedding} to
reconstruct the underlying dynamics of computer performance, then use
the resulting model to forecast the future values of computer
performance metrics such as memory or processor loads
\cite{josh-ida2011}.  In the case of simple microkernels like the one
that produced the trace in Figure~\ref{fig:ipc}, this deterministic
modeling and forecast strategy works very well.  In more-complicated
programs, however, such as speech recognition software or compilers,
this forecast strategy---as well as the traditional methods---break
down quickly.

This paper is a first step in understanding when, why, and how
deterministic forecast strategies fail when they are applied to
deterministic systems.  We focus here on the specific example of
computer performance.  We conjecture that the complexity of traces
from these systems---which results from the inherent dimension,
non-linearity, and non-stationarity of the dynamics, as well as from
measurement issues like noise, aggregation, and finite data
length---can make those deterministic signals \emph{effectively}
unpredictable.  We argue that \emph{permutation entropy}
\cite{bandt2002per}, a method for measuring the entropy of a
real-valued-finite-length time series through ordinal analysis, is an
effective way to explore that conjecture.  We study four
examples---two simple microkernels and two complex programs from the
SPEC benchmark suite---running on different Intel-based machines.  For
each program, we calculate the permutation entropy of the processor
load (instructions per cycle) and memory-use efficiency (cache-miss
rates), then compare that to the prediction accuracy attainable for
that trace using a simple deterministic model.

@@

To Add to this section:

\begin{enumerate}
\item Description of ARIMA
\item Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.

\item Maybe add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the begining of the results.

\end{enumerate}

@@

This is an extremely powerful result: among other things, it means
that one can build a formal model of the full system dynamics without
measuring (or even knowing) every one of its state variables.  This is
the foundation of the modeling approach that is used in this paper.
The first step in the process is to estimate values for the two free
parameters in the delay-coordinate map: the delay $\tau$ and the
dimension $m$.  We follow standard procedures for this, choosing the
first minimum in the average mutual information as an estimate of
$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
method of \cite{KBA92}, with a threshold of 10\%, to estimate $m$.  A
plot of the data from Figure~\ref{fig:ipc}, embedded following this
procedure, is shown in Figure~\ref{fig:embedding}.

 \begin{figure}
   \centering
     \includegraphics[width=\textwidth]{figs/colipc3d.png}
     \caption{A 3D projection of a delay-coordinate embedding of the trace
 from Figure~\ref{fig:ipc} with a delay ($\tau$) of 100,000 instructions.
 }
 \label{fig:embedding}
 \end{figure}

@@

Given a nonlinear model of a deterministic dynamical system in the
form of a delay-coordinate embedding like Figure~\ref{fig:embedding},
one can build deterministic forecast algorithms by capturing and
exploiting the geometry of the embedding.  Many techniques have been
developed by the dynamical systems community for this purpose
(e.g.,~\cite{casdagli-eubank92,weigend-book}).  Perhaps the most straightforward
is the ``Lorenz method of analogues'' (LMA), which is essentially
nearest-neighbor prediction in the embedded state
space~\cite{lorenz-analogues}.  Even this simple algorithm---which
builds predictions by finding the nearest neighbor in the embedded
space of the given point, then taking that neighbor's path as the
forecast---works quite well on the trace in Figure~\ref{fig:ipc}, as
shown in Figure~\ref{fig:cachePredTS}.
%
\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/colCachePredTS}
    \caption{A forecast of the last 4,000 points of the signal in
      Figure~\ref{fig:ipc} using an LMA-based strategy on the
      embedding in Figure~\ref{fig:embedding}.  Red circles and blue
      $\times$s are the true and predicted values, respectively;
      vertical bars show where these values differ. }
\label{fig:cachePredTS}
\end{figure}
%
On the other hand, if we use the same approach to forecast the
processor load\footnote{Instructions per cycle, or IPC} of the {\tt
  482.sphinx3} program from the SPEC cpu2006 benchmark suite, running
on an Intel i7\textsuperscript{\textregistered}-based machine, the
prediction is far less accurate; see Figure~\ref{fig:predsphinx}.
%
\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/sphinxPredicTS}
     \caption{An LMA-based forecast of the last 4,000 points of a
       processor-load performance trace from the {\tt 482.sphinx3}
       benchmark.  Red circles and blue $\times$s are the true and
       predicted values, respectively; vertical bars show where these
       values differ.}
\label{fig:predsphinx}
\end{figure}

@@

The weighted permutation entropy for the {\tt SVD} program is given in
Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
the time series. Within each of those windows, the statistics over words of
length 4 are computed and the WPE is calculated. The gray bands denote regions
where the 5,000 value window overlapped visually-distinct regimes. It can be
seen that the behaviors of the weighted permutation entropy vary between
regimes. [[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series different levels of complexity and predictability exist. Maybe point at some of the predicting predictability papers.]]

%\begin{figure}[htbp]
%  \centering
%  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
%  \caption{[[Joshua: I think adding the colored SVD trace to this would be %good or putting it above this figure]]The weighted permutation entropy of %one run of SVD. The gray bands
%    are regions where the window overlaps regimes. The window size used is
%    $5,000 \times 100,000$ instructions and the word length is $4$.}
%  \label{fig:wwpe}
%\end{figure}

%%FIGURE INTENT
%This figure shows that entropy changs over time with the signal as well as why the regimes for \svd were chosen. 
%%%%%%%%%%%%%%%%%
\begin{figure}[htbp]
  \centering
         \caption{
[Joshua: I think adding the colored SVD trace to this would be good or putting it above this figure but need to figure how to line them up properly. Also we need to label that the numbers on the bottom of WPE are regimes not instructions...]]The weighted permutation entropy of one run of SVD. The gray bands
    are regions where the window overlaps regimes. The window size used is
    $5,000 \times 100,000$ instructions and the word length is $4$.}\label{fig:wwpe}
  \begin{subfigure}{0.3\textwidth}
    \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
    \caption{The instructions per cycle of \svd. Each color corresonds to the different regimes as selected by rapid shifts in WPE, as seen in Figure \ref{fig:svd_wwpe}. From left to right each change in color represents a change in regime for 6 regimes in total. }
    \label{fig:svd_ts}
  \end{subfigure}%
  %\\
  \begin{subfigure}{0.3\textwidth}
    \includegraphics[width=0.3\textwidth]{figs/SVD_wwpe}
    \caption{The MASE of ARIMA vs weighted permutation entropy. }
    \label{fig:svd_wwpe}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   % \caption{The MASE values for LMA against ARIMA. The dashed line is the
   % identity, delineating the traces for which either LMA or ARIMA %performed
%    better. All traces except those from {\tt SVD$_1$} lie above the line,
%    indicating that LMA is better suited prediction method for the %computer
%   traces considered.}\label{fig:lma_vs_arima}
  %\end{subfigure}
\end{figure}

@@

%\item Introduce MASE and explain why random walk isn't *that* naive
%\item Paragraph on how to interpret MASE

@@

In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$. The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:compModel}. The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary. In particular, we perform $k$ 1-step predictions, after each 1-step prediction we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction. This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.}

As a figure of merit we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals: 
%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
The scaling term for MASE:
$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$. This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." The major advantage of MASE is that it allows fair comparison across methods, prediction horizons and varying signal scales. When a forecast results in a $MASE<1$ this means that the prediction method gave, on average, smaller errors than the 1-step errors from the in-sample random walk forecast strategy. Analogously, $MASE>1$ means that the prediction method did worse, on average than the 1-step errors for the in-sample random walk forecast strategy. In Table \ref{tab:error} we provide the distribution [[Joshua: Ryan, Is this the right word? we give mean $\pm$ std. dev but some have very skewed right tails]]  of MASEs for each of the 8 signals and 3 prediction strategies, these are averaged over 15 runs of each type (signal + method). For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for work lengths of $l=5$ and $l=6$.

\begin{table}[htdp]
\caption{MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal.[[Joshua:Maybe delete $l=5$ as we don't use it in any figures]] }
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve & $l=5$  & $l=6$ \\
\hline
\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $1.7970\pm0.0095$& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\

@@

%\item First paragraph: WPE is a good measure of predictability.  Figure:
%best athlete MASE vs. its WPE.

As can be seen in Table \ref{tab:error} the relationship between prediction accuracy and the weighted permutation entropy (WPE) is much as we conjectured: performance traces with high WPE are
indeed harder to predict using the forecasting models described in Section \ref{sec:compModel}. Figure~\ref{fig:pred_vs_ent} demonstrates this primary finding, with
the exception of one outlier whose behavior we will explain. In Figure~\ref{fig:pred_vs_ent} we plot the best prediction (i.e., the lowest MASE over all 3 methods over all runs of that program) for each of the 8 signals.  
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
  \caption{The best MASE among all runs and prediction methods vs weighted
  permutation entropy. For each of these, the word length used is $6$. The
  dashed line is a least-squares linear fit of all the points except for {\tt
  SVD$_1$} which we have excluded for reasons explained in the text.}
  \label{fig:pred_vs_ent}
\end{figure}
We find that the relationship between the two is roughly linear. The single outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE: namely
that in the absence of any large features, as all the other regimes have, the
WPE effectively falls back to the standard PE for which the noisy behavior
drives toward 1.0.

It may be tempting to paraphrase this primary finding as: ``Signals that are more complex are harder to predict.", but this truly misses the impact of this work. A more correct statement would be ``Time series which exhibit structure-less complexity are harder to predict for any model which uses structure as it's information processor. " [[I don't like this at all]]

\begin{itemize}
\item Complexity need not be hard to predict (can point at the simple predictions paper)
\item random walk for example is best predicted by guess what just happend
\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not. 
\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this.
\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.

\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured. And a constant, linear and nonlinear prediction strategy all fail. We should be able to conclude that guessing random values is the best we can do as is shown by MASE

\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction

\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method. That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy. This gives a practitioner leverage on when to give up and when to keep working.

\end{itemize}

Of course a complex signal is harder to predict than a trivial one, but this is \emph{not} the core contribution of this work as we will elucidate with Figure \ref{fig:lma_vs_arima}. [[Joshua: Make sure we make it clear WPE not only tells you when there is strucure to use but also what kinds of complexity are usable. e.g., white noise and deterministic chaos are both complex but one is predictable. The kind of complexity is important. Talk about chaos and random walk being complex but from a predictive standpoint that complexity spectrum is important **AND** quantifiable by WPE ]]
\begin{figure}[htbp]
  \centering
       \caption{
For each of
    these, the word length used is $6$.       
       The MASE values for LMA against ARIMA. The dashed line is the
    identity, delineating the traces for which either LMA or ARIMA performed
   better. All traces except those from {\tt SVD$_1$} lie above the line,
  indicating that LMA is better suited prediction method for the %computer
   traces considered.}\label{fig:lma_vs_arima}  
  \begin{subfigure}{\textwidth}
    \includegraphics[width=1.0\textwidth]{figs/LMA_prediction_vs_entropy}
    \caption{The MASE of LMA vs weighted permutation entropy. }
    \label{fig:lma_pred_vs_ent}
  \end{subfigure}%
  \\
  \begin{subfigure}{\textwidth}
    \includegraphics[width=1.0\textwidth]{figs/ARIMA_prediction_vs_entropy}
    \caption{The MASE of ARIMA vs weighted permutation entropy. }
    \label{fig:arima_pred_vs_ent}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   % \caption{The MASE values for LMA against ARIMA. The dashed line is the
   % identity, delineating the traces for which either LMA or ARIMA %performed
%    better. All traces except those from {\tt SVD$_1$} lie above the line,
%    indicating that LMA is better suited prediction method for the %computer
%   traces considered.}\label{fig:lma_vs_arima}

  %\end{subfigure}
\end{figure} 
In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly
compare the performance of the LMA and ARIMA prediction methods (respectively)
to the value of the weighted permutation entropy for all runs of each program
under consideration. The LMA MASE values are largely similar to those of the
best predictions, primarily because LMA often performed superior to ARIMA (and
the na\"ive method). So it is not surprising that the MASE from LMA and the corresponding MASE are highly correlated. On the other hand, the ARIMA MASE values are largely
uncorrelated with WPE values.
